---
title: "1.1 Getting started"
subtitle: "PLSC30500, Fall 2021"
# author: "Andy Eggers & Molly Offer-Westort"
output: 
  xaringan::moon_reader:
    css: [default, shinobi]
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
  
---


## This is it 

$$ \frac{\text{cov}(Y, X)}{\text{var}(X)}$$

---
class: inverse, middle, center 

## What do you love? 

I love **regression**.

.pull-left[
on your left
]

.pull-right[
and on your right
]


---

background-image: url("http://bit.ly/cs631-donkey")
background-size: cover

# Would you ever want to have an animated background? 


---

# Maybe not, but you would want this 

```{r echo = F}
knitr::include_graphics("http://bit.ly/cs631-pygmy")
```

---

# Code highlighting

```{r example-plot, fig.height = 3, fig.align = "center"}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<

```

---

# Comments on slides

If you press "p", you get to see the comments for this slide.

???

Isn't that interesting?

---

## The sky is the limit



