---
title: "Lecture 2.1 in-class code and follow-up"
author: "Andy Eggers"
date: "10/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

We used this code in class to show code completion: 

```{r}
dat <- read_csv(file = "./../data/vdem_2010_extract.csv")

dat %>% 
  select(country_name, continent_name)

```
## Recoding variables

I wrote on the board how to use `case_when` to create a new variable on the basis of one or more existing variables.

First we show the varaible we want to recode:

```{r}
vd <- vdemdata::vdem

vd %>% 
  count(e_regionpol_6C)
```

Now we create a new variable to replace the numerical regions with text strings:

```{r}
vd %>% 
  mutate(region = case_when(
    e_regionpol_6C == 1 ~ "E. Europe & Central Asia",
    e_regionpol_6C == 2 ~ "Latin America & Caribbean",
    e_regionpol_6C == 3 ~ "Middle East & N. Africa",
    e_regionpol_6C == 4 ~ "Sub-Saharan Africa",
    e_regionpol_6C == 5 ~ "W. Europe & N. America",
    e_regionpol_6C == 6 ~ "Asia & Pacific")) %>% 
  count(region)
```

Another option suggested in class is `recode`. (I said I don't use this but I realized later that I actually do!)

```{r}
# demonstrating recode for recoding 
vd1 <- vd %>% 
  mutate(region = recode(e_regionpol_6C, 
                         `1` = "E. Europe & Central Asia", 
                         `2` = "Latin America & Caribbean", 
                         `3` = "Middle East & N. Africa", 
                         `4` = "Sub-Saharan Africa", 
                         `5` = "W. Europe & N. America", 
                         `6` = "Asia & Pacific")) 

vd1 %>% 
  count(region)
```

We stored the result of that recode in `vd1` so now we can use it to reproduce the plot we made in class.

```{r}
# plotting year-region average polyarchy over time
vd1 %>% 
  group_by(year, region) %>% 
  summarize(mean_polyarchy = mean(v2x_polyarchy, na.rm = T)) %>% 
  ggplot(aes(x = year, y = mean_polyarchy, col = region)) + 
  geom_line()
```

At the end of lecture I showed how you can get something similar using `geom_smooth()`. You can get closer if you specify method `loess` and reduce the default `span` argument. (I also drop the confidence intervals using `se = F` and make the lines thinner to match the default for `geom_line()`.) It's a little bit smoother than just the grouped means, but if I make the span smaller I get an error.

```{r, warning = F, message = F}
# demonstrating using geom_smooth to get something very close to group means 
vd1 %>% 
  ggplot(aes(x = year, y = v2x_polyarchy, col = region)) + 
  geom_smooth(method = "loess", span = .02, se = F, lwd = .5)
# by trial and error I set the span as small as I could.
```


